Homework #6 
============================================================
_Huiting Ma_ 
 
 
**This homework will focus on the following parts:**
* <div id="Introduction1">[Introduction](#Introduction2)
* <div id="Data Cleaning1">[Data Cleaning](#Data Cleaning2)
* <div id="Data Aggregation and Visulation1">[Data Aggregation and Visulation](#Data Aggregation and Visulation2)
* <div id="Conclusion1">[Conclusion](#Conclusion2)
  
```{r include = FALSE}
opts_chunk$set(tidy = FALSE)
```

### <div id="Introduction2">[Introduction](#Introduction1)
This homework is to analyze the **Natural Disasters** around the world from the year 2000 to 2013. The natural disasters include earthquake, volcano and mass movement. My main focus is to identify which continents or countries have more natural disasters and whether the number of death, injuries, homeless and the population affected are correlated with the damaged amount. The dataset I am going to use is from EM-DAT, which is the International Disaster Database [here](http://www.emdat.be/database).

The definition of all variables can be found in the above website, which are:

- `NumDisaster` A unique disaster number for each event 
- `Country` Country (ies) in which the disaster has occurred
- `Year` When the disaster occurred. 
- `NumKilled` Persons confirmed as dead and persons missing and presumed dead (official figures when available)
- `NumInjured` People suffering from physical injuries, trauma or an illness requiring medical treatment as a direct result of a disaster
- `NumHomeless` People needing immediate assistance for shelter
- `NumAffected` People requiring immediate assistance during a period of emergency; it can also include displaced or evacuated people
- `TotalDamUSD` Several institutions have developed methodologies to quantify these losses in their specific domain. However, there is no standard procedure to determine a global figure for economic impact. Estimated damages are given (000') US$

### <div id="Data Cleaning2">[Data Cleaning](#Data Cleaning1)

```{r}
library(plyr)
library(xtable)
library(ggplot2)
library(reshape2)
```

Supercheck whether data has imported correctly
```{r}
NaturalDisaster <- read.csv("disaster.csv") 
str(NaturalDisaster)
```

Learn from Jennie, to define a function for converting and printing to HTML table.
```{r}
htmlPrint <- function(x, ...,
                      digits = 0, include.rownames = FALSE) {
  print(xtable(x, digits = digits, ...), type = 'html',
        include.rownames = include.rownames, ...)
}
```

Try to count the number of natural disaster over time on different continents
```{r results='asis'}
numCountByYear <- daply(NaturalDisaster,~Year + 
                          Continent, summarize, 
                        TotalCount = sum(NumDisaster))
numCountByYear <- as.data.frame(numCountByYear)
htmlPrint(numCountByYear)
```

Seems 2013 does not have a lot of observations, Let's drop 2013.
```{r}
NDisaster <- droplevels(subset(NaturalDisaster,Year != "2013"))
table(NDisaster$Year) #Check whether 2013 has dropped
```

Try to count the number of countries in each continents
```{r results='asis'}
numCountries <- ddply(NDisaster, ~Continent, summarize, numCoutries = length(unique(Country)))
htmlPrint(numCountries)
```

Now, let us try to plot a graph and visualize the results
```{r}
ggplot(NaturalDisaster, aes(x = Year, y = NumDisaster, color = Year)) +
  geom_jitter() + facet_wrap(~ Continent) +
  ggtitle("How is Number Disasters Changing over Time on Diffferent Continents")
```
Based on the graph, it is found that "Oceania" does not have a lot of data, drop it!

```{r}
NDisaster <- droplevels(subset(NaturalDisaster,Continent != "Oceania"))
table(NDisaster$Continent) #Check whether "Oceania" has dropped
```

Also, I found there are some missing data for Number of affected. Let's drop these data!
```{r}
NDisaster <- droplevels(subset(NDisaster,NumAffected != "0"))
str(NDisaster)
```

Since there also a lot of missing data for Total damage, let's drop these values.
```{r}
Disaster <- droplevels(subset(NDisaster,TotalDamUSD != "0"))
str(Disaster)
```

Since I want to fit **linear regression model** in the second script, it is necessary to make sure
all countries should have multiple observations. I will delete the countries with only 2 or less
observations.

```{r}
NumObsCountry <- ddply(Disaster, ~Country, summarize, numobs = length(Year))
CountryDrop <- droplevels(subset(NumObsCountry,!(numobs > 2)))
Disaster <- droplevels(subset(Disaster,
                              !(Country %in% CountryDrop$Country)))
```

```{r results = 'asis'}
DropNumObsCountry <- ddply(Disaster, ~Country, summarize, numobs = length(Year))
htmlPrint(DropNumObsCountry)
```

Yes, countries with only few observations have been dropped. The last thing I want to do is 
to reorder continents based on the total nunmber of disasters happened.

```{r}
Disaster <- within(Disaster, Continent <- reorder(Continent, NumDisaster, sum))
Disaster <- arrange(Disaster,Continent)
```

### <div id="Data Aggregation and Visulation2">[Data Aggregation and Visulation](#Data Aggregation and Visulation1)

In this stage, dataset has been sucessfully cleaned to the format that I want. Let us get into 
**Exploratory Data Analysis** stage.

**First, try to get the spread of death within the continents.**

```{r results='asis'}
spreaddeath <- ddply(Disaster, ~ Continent, summarize,
                     sdNumKilled = sd(NumKilled), 
                     iqrNumKilled = IQR(NumKilled))
htmlPrint(arrange(spreaddeath, sdNumKilled))
```

It can be seen that there are a lot of variation for the number of killed people in natural disasters.

```{r fig.width=10, fig.height=10}
newspread <- melt(spreaddeath, id="Continent")
ggplot(newspread, aes(x = Continent, y = value, colour = variable)) + 
     geom_point() + geom_line(aes(x=as.numeric(Continent)))+ ylab("spread") +
     ggtitle("Measure of Spread")
```

Then,try to produce the maximum and minimum statistics for 
different variables in all continents for all continents
```{r results='asis'}
maxmean<- ddply(Disaster, ~ Continent, summarize, 
                meanDamage = mean(TotalDamUSD),
                maxDamage= max(TotalDamUSD),
                meanAffPop = mean(NumAffected),
                maxAffPop = max(NumAffected),
                meanInjured = mean(NumInjured),
                maxInjured = max(NumInjured),
                meanDeath = mean(NumKilled),
                maxDeath = max(NumKilled),
                meanHomeless = mean(NumHomeless),
                maxHomeless = max(NumHomeless)
)
htmlPrint(arrange(maxmean,maxDamage))
```

We have summarized some detailed information about different 
variables in the dataset. Now, let's discouver more details.
First,let's identify which varibales play an important role to predict 
the damaged amount from natural disasters

To start with, Let us to fit the full model.
```{r}
FullModel <- lm(TotalDamUSD ~ NumDisaster + NumKilled + NumInjured +
                NumAffected +NumHomeless, data = Disaster)
summary(FullModel)
```

Based on the summmary, it can be shown that number of disasters 
played a significant role to determine the damaged amount. 
Therefore, I will pay more attention on whether
number of disasters also play an essential role in country level.

```{r}
mFun <- function(x) {
  model <- lm(TotalDamUSD ~ NumDisaster, x)
  estCoefs <- c(coef(model))
  # estSE <- c(se.coef(model))
  ## 2 means 2nd row (we do not want to test intercept), 
  ## 4 means the 4th column, which corresponding to p-value
  p_value <- summary(model)$coefficients[2, 4]  
  names(estCoefs) <- c("intercept", "NumDisaster")
  #names(estSE) <- c("SE(intercept)", "SE(NumDisaster)")
  names(p_value) <- "p-value (NumDisaster)"
  return (c(estCoefs,p_value))
}

mCoefs <- ddply(Disaster, ~Country, mFun)
print(mCoefs)
```
**Question:** For this part, I did not use `htmlPrint()` since I do not know how to control
decimals in r markdown. R markdown automatically round numbers.

I tried my best, but I am not sure whether this is the correct way.
Based on above table, we found that even though the number 
of disasters play an critical role in the world level, 
it does not significant in country level for most countries. 
It might because the individual country has limited sample size. 
Thus, the standard error is huge.


**The next topic that I would like to focus one is how is the number of disasters changing over 
time on different continents.To begin with, let us plot the number of disasters in different continents.**

To begin with, let us plot the number of disasters in different continents.


```{r fig.width=10, fig.height=10}
ggplot(Disaster,aes(x=NumDisaster, fill= Continent)) + facet_wrap(~Continent)+
  geom_bar(binwidth = 2, color = "black") +
  ggtitle("Number of Disasters in Diffferent Continents")
```

After having some basic ideas about the frequency of the number of disasters in 
different continents,I will look at the number of disasters changing over time 
on diffferent continents

```{r fig.width=10, fig.height=10}
p <- ggplot(Disaster, aes(x = Year, y = NumDisaster, color = Year))+ 
  geom_jitter() + facet_wrap(~ Continent) + 
  geom_line(stat = "summary", fun.y = "mean", col = "red", lwd = 1) +
  ggtitle("How is Number of Disasters Changing over Time on Diffferent Continents") + 
  scale_x_continuous(name = "Year", breaks = seq(min(Disaster$Year), 
                                                 max(Disaster$Year), by = 2))  + 
  xlab("Year") + ylab("Number of Disasters")
print(p)
```
Based on the plot, there is no significan relationship beween the number of disasters 
and time. However, it can be found that there is some difference for the number of 
disasters across continents.

Try boxplot for the year 2000, 2005 and 2010
```{r fig.width=10,fig.height=10}
ggplot(subset(Disaster, Year %in% c(2000,2005,2010)), aes(x = factor(Year), 
                                                          y = NumDisaster, 
                                                          fill = Continent), 
       groups = Continent) + geom_boxplot(alpha = 0.2, outlier.colour= "red") 
```
Obviously, Asia has more variation for the number of disasters.

Finally, let us try another plot about the number of disasters changing over 
time on diffferent continents
```{r fig.width=10, fig.height=10}
low_number= 10
ggplot(Disaster, aes(x = Year, y = NumDisaster, 
                        colour = NumDisaster <= low_number)) + 
  geom_jitter(position = position_jitter(width = .2)) + 
  facet_wrap(~ Continent) + 
  ggtitle(paste("NumDisaster <= ", low_number)) + 
  theme(plot.title = element_text(face="bold")) + 
  scale_colour_discrete(name="",breaks=c("FALSE", "TRUE"),
                        labels=c("Death > 5", "Death <= 5"))+
  ggtitle("Show Number of Disasters by Year across Continents")
```


**Now,Let us look at a special plot: dots scatterplot of number of natural disaster over year for China. Why?**

```{r fig.width=10, fig.height=10}
ggplot(subset(Disaster, Country == "China P Rep"), aes(x = Year, y = NumKilled)) + 
  geom_line() + xlab("Year") + ylab("Number killed") +
  ggtitle("How is Number of People Killed over time in People's Repubic of China") +
  scale_x_continuous(name = "Year", breaks = seq(min(NDisaster$Year), 
                                                 max(NDisaster$Year), by = 2)) 
```

The graph shows that more than 80,000 were killed by natural disaster 
in the year of 2008 in China. That was Sichuan Earthquake. At that time, 
I was also in Sichuan and experienced this catastrophe. I feel sorry for 
those people who lost their lives during this disaster.Since I experienced 
this natural disaster, I would think this is an important plot for me!!

**Next, I would like to evaluate whether there is a relationship between affected population and death across continents.**

The question is that whether more people got affected by natural disasters will lead more people be killed? Since some natural disasters caused a lot of people lost lives, I used log transformation (**Question** Log transformation may not make sense?)

```{r fig.width=10, fig.height=10}
ggplot(Disaster, aes(x = NumAffected, y = NumKilled, color = Continent)) + 
  geom_point() + scale_x_log10() + scale_y_log10()+
  ggtitle("How NumKilled related to NumAffected across Continents") 
```
Based on the graph, it can be seen that with the increase of number of people got affected,
there is a slightly increasing trend for the number of people got killed.

### <div id="Conclusion2">[Conclusion](#Conclusion1)
In conclusion, the number of disasters played a significant role to determine the total damaged amount in population level but not in country level. In addition, with the increase of number of people got affected,
there is a slightly increasing trend for the number of people got killed.